Random Matrix Theory Tools for the Analysis of Functional Magnetic Resonance Imaging Examinations

### 2022-10-31 paper/paper/SPIE/article.tex
[SHARE]

\section*{Purpose}
Random matrix theory (RMT) is an increasingly useful tool for understanding large, complex systems.
Prior studies have examined functional Magnetic Resonance Imaging (fMRI) images using tools from
RMT, with some success. However, RMT computations are complex, and sensitive to numerous analytic
choices, and previous investigations have been mostly explanatory / descriptive. We systematically
investigate the usefulness of RMT in a variety of fMRI datasets in a predictive framework.

@@

\section*{Results}
We find eiegenfeatures to be broadly predictive in a wide variety of contexts, and that RMT-specific
features like the spectral rigidity and level number variance have considerable potential in
feature-engineering for fMRI. On median, we find ...

\section*{Conclusions}
Use RMT yo.

@@

\section{Introduction}
\label{sect:intro}  % \label{} allows reference to this section
This document shows the format and appearance of a manuscript prepared for submission to an SPIE journal. Note that this template is only intended to be used as a guideline for author convenience. It is designed for optimum clarity and ease of reading for editors and reviewers, but the template does not reflect the final page layout of a published journal paper. Accepted papers are professionally typeset in XML according to the layout and design of the journal.

\subsection{Use of This Document}

This document is prepared using LaTeX2e\cite{Lamport94,Goossens97} with the class file {\ttfamily spieman.cls}.  The LaTeX source file used to create this document is {\ttfamily article.tex}, which contains important formatting information embedded in it. Authors may use it as a template to create their own manuscript. While LaTeX properly handles most formatting issues, the author may occasionally need to intervene to obtain a satisfactorily  formatted manuscript.

\subsection{English}

Authors are strongly encouraged to follow the principles of sound technical writing, as found in Refs.~\citenum{Alred03} and \citenum{Perelman97}, for example. In addition, good English usage is essential. Authors whose native language is not English may wish to collaborate with a colleague whose English skills are more advanced. Alternatively, you may wish to have your manuscript professionally edited prior to submission by Editage, our recommended independent editorial service: \linkable{https://www.editage.com/spie/}. SPIE authors will receive a 15\% discount off their services. A spell checker can be helpful to discover misspelled words, but authors should also proofread their papers carefully prior to submission. Manuscripts that do not meet acceptable English standards or lack clarity may be rejected.

>>>>

\section*{Purpose}
Random matrix theory (RMT) is an increasingly useful tool for understanding large, complex systems.
Prior studies have examined functional Magnetic Resonance Imaging (fMRI) scans using tools from
RMT, with some success. However, RMT computations are complex, and sensitive to numerous analytic
choices, and previous investigations have been mostly explanatory / descriptive. We systematically
investigate the usefulness of RMT in a variety of fMRI datasets in a predictive framework.

@@

\section*{Results}
We find eiegenfeatures to be broadly predictive in a wide variety of contexts, and that RMT-specific
features like the spectral rigidity and level number variance have considerable potential in
feature-engineering for fMRI. On median, we find ... However, we also find the utility of the
extracted eigenfeatures to depend strongly and complexly on various analytic decisions.

\section*{Conclusions}
Consider using RMT.

@@

\section{Introduction}
\label{sect:intro}  % \label{} allows reference to this section
% This document shows the format and appearance of a manuscript prepared for submission to an SPIE
% journal. Note that this template is only intended to be used as a guideline for author convenience.
% It is designed for optimum clarity and ease of reading for editors and reviewers, but the template
% does not reflect the final page layout of a published journal paper. Accepted papers are
% professionally typeset in XML according to the layout and design of the journal.

In functional magnetic resonance imaging (fMRI), changes in the blood-oxygenation-level-dependent
(BOLD) signals are related to neural activity. It is common to investigate statistical relationships
between the BOLD signals through \textbf{functional connectivity} analyses, where correlations between
collections of these signals are examined to infer connections between different voxels or regions
of interest (ROIs) within the brain.

Whether in the presence of experimental stimuli, or the relative absence, as in a resting state,
complex functional connectivity networks are ubiquitous
\citenum{bucknerBrainDefaultNetwork2008,foxCoverHumanBrain2005,gonzalez-castilloTaskbasedDynamicFunctional2018,hermundstadStructuralFoundationsRestingstate2013}.
This complexity suggests fMRI is a candidate to be studied using \textbf{Random Matrix Theory} (RMT), a set
of mathematical tools originally developed some 50 years ago to solve complex problems in nuclear
physics \citenum{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004}. In various
physical systems where nuclei can have different possible states, precisely modeling the entirety of
the interactions between nuclei is often intractable. For such large, complex systems, the
eigenvalues (or spectra) efficiently summarize the totality of the interactions between components
of the system, and RMT describes the expected behaviour of such eigenvalues. By analysing the
spectrum statistically, complex systems can be investigated by comparison to the universal
properties predicted by RMT.

In recent years, RMT analyses have been applied to study brain functioning.  The earliest  study
demonstrated that spectra of the correlations between electroencephalographic (EEG) signals closely
resemble those of the Gaussian Orthogonal Ensemble (GOE) \citenum{sebaRandomMatrixAnalysis2003}.
Somewhat more recently, RMT was used to evaluate the quality of whole brain features extracted from
fMRI data \citenum{voultsidouFeatureEvaluationFMRI2007,verganiRestingStateFMRI2019}. RMT has also
been used in diffusion MRI to aid in the selection of the number of components to employ in
principal-component reduction analysis and denoising
\citenum{veraartDenoisingDiffusionMRI2016,verganiRestingStateFMRI2019,ulfarssonDimensionEstimationNoisy2008}.

Finally, RMT has been used in ROI-based fMRI functional-connectivity studies to investigate
differences between rest and task states \citenum{wangSpectralPropertiesTemporal2015}, between
subjects with and without ADHD \citenum{wangRandomMatrixTheory2016}, and between pain and non-pain
states \citenum{matharooSpontaneousBackpainAlters2020}. Across these three studies, the spectra of
resting or low-attention states exhibited properties closest to the GOE. These findings suggest that
certain aspects of psychological processes might be characterized, in part, by features computed
from the eigenvalues of fMRI correlation matrices, and that these features might vary in an
interpretable manner across psychological processes. If this is the case, RMT could have unique
potential in characterizing the functioning of the human brain.

\section{Datasets}

As our primary intention is to examine the potential utility of RMT-derived features across a wide
variety of data, and not to decisively put forward or refute any specific hypotheses relating RMT to
psychological processes, we examine a wide selection of fMRI datasets available on the
\href{https://openneuro.org/}{OpenNeuro} platform [36].  This allows for a broad assessment of the
potential of voxelwise RMT features.

However, since previous studies employing RMT for feature extraction have partially interpreted
their findings with respect to attentional processes
\citenum{wangRandomMatrixTheory2016,matharooSpontaneousBackpainAlters2020}, we have elected to use
datasets that involve attention in some way. This includes measures of attention or vigilance, or
comparisons between a condition involving more attention and focus (e.g. a task state) or conditions
which should involve considerably less attention or focus (e.g. resting-state).

Dataset scan parameters and acquisition details are summarized in Table \ref{table:1}, and sample
and subgroup sizes are summarized in Table \ref{table:2}. For all datasets, we use the raw
(unpreprocessed) scans, and perform our own preprocessing (detailed later). Given the complexity of
these datasets, we highlight below only those aspects relevant to the current paper.

\begin{table}[h!]
\small
\centering
\begin{tabular}{ c c c c c c c }
\hline
\textbf{ID}    & \textbf{Dimensions}  & \textbf{Voxel size (mm)} & \textbf{TR} & \textbf{Volumes} \\
\hline
PSYCH     & 128 x 128 x 70 & 1.5 x 1.5 x 1.5 & 3.0  & 300  \\
PARK      & 80 x 80 x 43   & 3.0 x 3.0 x 3.0 & 2.4  & 149  \\
OSTEO     & 64 x 64 x 36   & 3.4 x 3.4 x 3.0 & 2.5  & 300  \\
BILING    & 100 x 100 x 72 & 1.8 x 1.8 x 1.8 & 0.88 & 823  \\
LEARN     & 64 x 64 x 36   & 3.0 x 3.0 x 3.0 & 2.0  & 195  \\
DEPRESS   & 112 x 112 x 25 & 2.0 x 2.0 x 5.0 & 2.5 & 100   \\
AGING     & 74 x 74 x 32   & 3.0 x 3.0 x 4.0 & 2.0 & 300   \\
\hline
\end{tabular}
\caption{
    ID = Identifier for paper. TR = Time of Repetition (seconds).
    Time = total duration (minutes) of each scan. Dimensions listed as M x N x P,
    indicate P slices with dimensions M x N.
}
\label{table:1}
\end{table}

\begin{table}[h!]
\small
\centering
\begin{tabular}{ l l c c c }
\hline
\textbf{ID}  & \textbf{Subgroup / Condition}  & \textbf{Subjects}  & \textbf{Task} & \textbf{Scans per Subject} \\
\hline
PSYCH   &  high\_vigilance    & 12 / 11 & resting-state                    & 2 \\
        &  low\_vigilance     & 10 / 11 & resting-state                    & 2 \\
        &  trait\_attend      & 12 / 15 & resting-state                    & 2 \\
        &  trait\_non-attend  & 10 / 7  & resting-state                    & 2 \\
        &  task\_attend       & 13 / 15 & resting-state                    & 2 \\
        &  task\_nonattend    &  9 / 7  & resting-state                    & 2 \\
\hline
LEARN   &  task              & 24      & learn image sketches & 16 \\
        &  rest              & 24      & resting-state                    & 2 \\
\hline
REFLECT &  control (no pain) & 20      & resting-state                    & 1 \\
        &  duloxetine        & 19      & resting-state                    & 1 \\
        &  placebo           & 37      & resting-state                    & 1 \\
        &  allpain (duloxetine + placebo) & 56         & resting-state    & 1 \\
\hline
PARK    &  Parkinson's       & 17      & ANT                              & 6 \\
        &  control           & 17      & ANT                              & 6 \\
\hline
ECHO    &  summed/interleave & 31      & resting-state                    & 1 \\
        &  summed/interleave & 31      & imagery/reflection               & 4 \\
\hline
\end{tabular}
\caption{
    ID = Identifier for paper. FOV = Field of View. TR = Time of Repetition (seconds).
    Time = total duration (minutes) of each scan. Dimensions listed as M x N x P,
    indicate P slices with dimensions M x N.
}
\label{table:2}
\end{table}

\section{Methods}
